Adversarial Deep Neural Networks Effectively Remove Nonlinear Batch Effects from Gene-Expression Data

### 2019-05-16 thesis.tex
Address some of Dr. Piccolo's comments

% Title / abstract page
\begin{titlingpage}
	\setcounter{page}{2}
	\phantomsection
	\addcontentsline{toc}{section}{Abstract}
	\begin{center}
		ABSTRACT\\
		\vspace{1em}
		\begin{adjustwidth}{+0.5in}{+0.5in}
			\begin{center}
				\thetitle\\
			\end{center}
		\end{adjustwidth}
		\vspace{1em}
		\theauthor\\
		Department of Biology, BYU\\
		Master of Science
	\end{center}
	% \begin{abstract}
		Gene-expression profiling enables researchers to quantify transcription levels in cells, thus providing insight into functional mechanisms of diseases and other biological processes.
		However, because of the high dimensionality of this data and the sensitivity of measuring equipment, expression data often contains unwanted confounding effects that can skew analysis.
		For example, collecting data in multiple runs causes nontrivial differences in the data (known as batch effects),
		known covariates that are not of interest to the study may have strong effects,
		and there may be large systemic effects when integrating multiple expression datasets.
		Additionally, many of these confounding effects represent higher-order interactions that may not be removable using existing techniques that identify linear patterns.
		We created Confounded to remove these effects from expression data.
		Confounded is an adversarial variational autoencoder that removes confounding effects while minimizing the amount of change to the input data.
		We tested the model on artificially constructed data and commonly used gene expression datasets and compared against other common batch adjustment algorithms.
		We also applied the model to remove cancer-type-specific signal from a pan-cancer expression dataset.
		Our software is publicly available at \url{https://github.com/jdayton3/Confounded}.
	% \end{abstract}
	\vfill
	\noindent
	Keywords: batch effects, batch correction, gene expression, transcriptomics, deep learning, adversarial neural network, variational autoencoder
\end{titlingpage}

@@

Gene expression data can be applied in many ways to advance our understanding of medicine and biology.
For example, expression data has been applied to discover conserved genetic modules \citep{stuart_gene-coexpression_2003}, to better understand the mechanisms of cardiovascular disease \citep{henriksen_application_2002}, to more accurately predict the clinical outcome in cancer patients \citep{veer_gene_2002}, and to discover effective drugs for treating specific diseases \citep{sirota_discovery_2011}.
Expression datasets are quite ``wide,'' often containing tens of thousands of columns representing each gene in the human transcriptome.
Because of the sensitive nature of this data (i.e. gene product levels in organisms may respond drastically to small environmental changes) and of the equipment used to quantify expression levels, expression data often contains unwanted confounding effects that can skew analysis.
Three examples of this include
\begin{enumerate*}[(a)]
	\item batch effects, \label{it:batch}
	\item known covariates, and \label{it:cov}
	\item dataset-level effects. \label{it:int}
\end{enumerate*}
\ref{it:batch} Batch effects occur when expression data are generated in multiple runs or multiple batches, and slight systemic differences occur during the different runs, such as different technicians operating the machine or slight temperature differences in the room.
Batch effects are understood to have a nontrivial impact on high-throughput expression data \citep{leek_tackling_2010}.
In one study, researchers found that, contrary to previous knowledge, expression values from mice and humans clustered more closely by species than by tissue type \citep{yue_comparative_2014};
however, referees showed in a rebuttal that when accounting for batch effects, these data actually clustered more closely by tissue type, as initially expected \citep{gilad_reanalysis_2015}.
\ref{it:cov} Systemic bias can be even more pronounced when there are known covariates---for example, using data collected from different cancer types when cancer type is not of interest to the study \citep{dayton_classifying_2017-1}.
\ref{it:int} Finally, though batch effects within a dataset collected all by one lab at one time may be somewhat decreased experimentally by careful replication of experimental conditions, this is no longer possible when integrating different datasets;
however, much better statistical significance can be achieved when analyzing larger datasets, so data integration is a key goal for transcriptomic analysis  \citep{lazar_batch_2013}.
Each of these problems---batch effects, known covariates, and dataset integration---represents a case where data measurements are effected by some hidden variables and must be removed for effective analysis (see Figure \ref{fig:workflow}).

@@

Several methods exist for removing batch effects from gene expression datasets.
Two commonly used methods are ComBat \citep{johnson_adjusting_2007} and SVA \citep{leek_capturing_2007}.
ComBat uses an empirical Bayes method to estimate batch effect parameters and then uses linear regression to remove the effects, and SVA uses singular value decomposition to model batch effects which can then be accounted for in statistical analyses.
Since both of these methods use linear methods to model batch effects, they are not designed to account for nonlinear effects, such as complex interactions in gene pathways in response to environmental changes.
As machine learning becomes more common in biological research, these nonlinear batch effects become more troublesome since many machine learning algorithms can successfully identify complex interactions between variables.
Advances in neural networks have introduced new ways to account for higher-order, nonlinear relationships in data.
These networks have proven effective in removing irrelevant, domain-specific signal in credit rating, online reviews, and image recognition tasks \citep{louizos_variational_2015}.
Recent work \cite{shaham_removal_2017,shaham_batch_2018,upadhyay_removal_2019} has applied neural networks to batch effects.
However, these recent programs have several items that complicate their usability on many real-world datasets:
they require that the input data only contains two batches, that the batches are sufficiently large, and that the batches are perfectly balanced.
These requirements rarely hold in existing datasets; for example, the bladderbatch dataset used in the documentation for the R sva package \cite{leek_bladderbatch_2017,leek_sva_2017} has 5 batches, only 57 samples, and between 4 and 19 samples per batch.
Additionally, the metrics these studies have used to validate results don't test whether complex interactions still remain in the data.

@@

In this study, we present Confounded, an adversarial autoencoder that identifies and removes effects between different batches.
We test the hypothesis that using an adversarial neural network can correct for batch effects more completely than previous tools do.
We also explore the extent to which batch effects still remain in different datasets after adjustment with various algorithms, and we present a framework to assess the extent to which batch effects remain after adjustment using various classification algorithms.

@@

We used an adversarial autoencoder network to model and remove the batch effects.
We structured this network in two parts: a variational autoencoder \cite{louizos_variational_2015} to replicate the input (expression) data and a discriminator to detect remaining batch effects in the autoencoder's output.
By penalizing the autoencoder for the discriminator's success, the autoencoder subnetwork learned over the course of training to output the expression data free of batch effects.
We implemented the neural network in TensorFlow 1.11.0 \cite{abadi_tensorflow_2015} with Python 3.6 \cite{python_software_foundation_python_2019}.
All layers in the network were fully connected and all activation functions were Rectified Linear Units (ReLUs) \cite{agarap_deep_2018} except the final layers in the autoencoder and the discriminator, which used the sigmoid function.

@@

We implemented the variational autoencoder \cite{louizos_variational_2015} described by \citet[Chapter 15]{geron_hands-machine_2017}.
Additional information can be found in our supplementary material.

@@

We trained the network using three loss functions.
First, we calculated the autoencoder's loss ($L_A$) by summing the reconstruction loss (sigmoid cross entropy between the autoencoder's input and output) and the latent loss (Kullback-Leibler divergence \cite{kullback_information_1951} of the code layer).
Second, we calculated the discriminator's loss ($L_D$) as sigmoid cross entropy between its output and a one-hot encoding of the samples' batch labels.
Finally, we also trained the autoencoder layers on a combination of the two previous losses,

@@

The $\lambda$ value represents a tradeoff parameter for tuning the network's tendency for more faithfully replicating the input or for more completely removing batch effects.
A higher $\lambda$ value indicates that the network should remove batch effects more aggressively, whereas a lower value indicates that the network should instead favor faithfully reconstructing the input data.
We did not optimize $L_A$ directly; instead we trained the autoencoder by optimizing $L_{dual}$.

@@

In all cases, we trained the network using the Adam Optimizer \cite{kingma_adam_2014} with a training rate of 0.0001 for 10,000 iterations on mini-batches of size 100.
In each iteration, we optimized on both $L_D$ and $L_{dual}$.
When optimizing $L_D$ (i.e. training the discriminator), we froze the autoencoder's weights, and vice versa.
We trained the network on a 2017 Dell XPS 15 9560 with an 8-core Intel i7-7700HQ CPU and 16 GB of RAM.
For each dataset, training typically took roughly 30 minutes to complete, including the time taken to load the input into memory and to save the output to disk.

\begin{figure}
	\pdfbookmark[1]{Figure \ref{fig:training_loss}}{training_loss}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/training_loss.pdf}
	\caption[Autoencoder and discriminator loss]{\textbf{Autoencoder and discriminator loss} over time for one run of Confounded.
	Over the course of training, the autoencoder more faithfully replicates the input data.
	The autoencoder also seems to introduce noise (see the red dashed line around iteration 3100) in response to the discriminator's slight improvements.}
	\label{fig:training_loss}
\end{figure}

@@

Because there is no batch information in the MNIST digits dataset, we had to simulate nonlinear batch effects.
To do so, we wrote a Python script to take the MNIST data in, apply a nonlinear effect, and output the adjusted data.
We applied nonlinear effects by iteratively realizing vectors of normally distributed values, multiplying and adding these vectors to the ``expression'' vectors, and applying nonlinearity to the adjusted vector.
We split the image data into two batches while keeping the batches balanced (5,000 images for each batch) and including the same number of each digit in each batch.
We applied the same random vectors to each image in a batch.
Finally, we added random noise to each image in order to prevent images in a batch from being overly similar to each other.

@@

The GSE37199 dataset contains Affymetrix microarray gene-expression data from patients with advanced castration-resistant prostate cancer \cite{olmos_prognostic_2012}.
We accessed a version of this data from \url{http://doi.org/10.17605/OSF.IO/SSK3T} that was tidied as part of a curated compendium of human transcriptional biomarker data \cite{golightly_curated_2018}.
It contains expression values for 93 tissue samples categorized as either ``Advanced castration resistant'' or ``good prognosis.''
We used this cancer status variable as the ``true class'' for classification.
It has two types of batch variables: ``plate'' and ``centre.''
We adjusted against the ``plate'' variable because it was more balanced than ``centre'' (with counts of $\{43, 50\}$ compared to $\{27, 66\}$).
The GSE37199 dataset represents a slightly larger dataset than bladderbatch, with only two batches that are closer to being balanced (with batch counts of $\{4, 5, 11, 18, 19\}$ and $\{43, 50\}$, respectively).

@@

We compared our method to two other batch adjusters: a mean and scale adjuster and ComBat \citep{johnson_adjusting_2007}.
We implemented the mean and scale adjuster in the R programming language, version 3.6.0 \citep{r_core_team_r_2014} using RStudio version 1.2.1194 \citep{rstudio_team_rstudio_2018}.
We used the ComBat implementation from the sva package \citep{leek_sva_2017} with some modifications to allow it to work on columns without variance in the MNIST dataset.
There are a number of other methods for batch adjustment that do and do not use deep neural networks (for example, see \citep{leek_capturing_2007,espin-perez_comparison_2018,shaham_removal_2017,shaham_batch_2018}). % TODO: add like 8 more methods
Unfortunately, most of these methods lack a common interface and common assumptions that input datasets must meet.
For these reasons, we compared Confounded only to the two methods listed above.
Future batch adjustment research may benefit from standardization of input formats, user interfaces, and validation datasets.

@@

\subsection{Confounded removes nonlinear batch effects that other adjusters miss}

Various previous studies use differing methods to show qualitatively and quantitatively how well batch effects have been removed.
To compare Confounded to other batch adjustment methods, we compared PCA and t-SNE plots along with MSE, MMD, and classifier batch prediction accuracy (using various classifiers).

@@

Confounded shows mixed success with the MSE and MMD metrics.
With MSE, Confounded outperformed the scale adjuster in 3 of the 4 datasets but scored drastically worse on the fourth dataset, with score pairs of \{(0.003, 0.00168), (0.247, 0.0698), (0.00312, 0.0187), (912315, 139458)\}.
With MMD, Confounded outperformed the scale adjuster again in 3 of the 4 datasets and tied the scale adjuster on the fourth dataset, with scores of \{(0.463, 0.258),(0.0906, 0.0376),(0.0653, 0.0117),(0.0942, 0.0942)\}.
With both metrics, Confounded consistently performed somewhat worse than ComBat.
(See Figures \ref{fig:mse} and \ref{fig:mmd}.)

@@

We would expect that after batch adjustment by an ideal adjuster, batch would no longer be detectable by any machine learning classifier.
Using the batch classification accuracy metric, Confounded seems to outperform other adjusters on larger datasets, whereas ComBat and Confounded seem to perform about the same on smaller datasets (see Figure \ref{fig:batch}).
With both the bladderbatch and GSE37199 datasets, batch classification accuracy decreases well below baseline after batch adjustment with ComBat for all classifiers we tested (see Table \ref{tab:batch}).
Interestingly, batch accuracy also decreases drastically for the MNIST and TCGA datasets, but only for the Naive Bayes classifier.
This may be due to two factors: both ComBat and Naive Bayes use Bayesian methods, so ComBat may specifically remove the effects that Naive Bayes identifies; and Naive Bayes does not find patterns based on interactions between variables.
Although Naive Bayes is no longer able to identify batch effects in the data after ComBat-adjustment, Random Forests (which does use interactions between variables) still has a very high accuracy for MNIST and an increased accuracy for TCGA.
In contrast, after adjustment by Confounded, the Random Forests algorithm's accuracy decreases more than with any other adjuster for both the MNIST and TCGA datasets.
This indicates that while ComBat's performance may work at least as well as Confounded for smaller expression datasets, Confounded may work better with larger datasets.

@@

With the larger datasets, Confounded's true class accuracy consistently decreases below the accuracy of other adjusters.
A look at the MNIST digits before and after adjustment (see Figure \ref{fig:mnist}) shows that Confounded's output is often blurry, as is common with the output of variational autoencoders \cite{hou_deep_2016}.
With MNIST, Confounded's accuracy with Random Forests is still much higher than baseline (84.8\% vs. 11.4\%), but with TCGA, the accuracy decreases below baseline (66.5\% vs. 69.8\%) while the other adjusters' accuracies remain above baseline.
However, preliminary testing indicates that fine-tuning Confounded's parameters to run on these datasets rather than using the default settings may greatly improve Confounded's post-adjustment true class accuracy.
% TODO: include this^

\begin{figure}
	\pdfbookmark[1]{Figure \ref{fig:true_class}}{true_class}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/true_class_accuracy.pdf}
	\caption[True class classification accuracy]{\textbf{True class classification accuracy} for several datasets and adjusters with 4-fold cross-validation repeated 3 times.
	See also Table \ref{tab:true_class}.}
	\label{fig:true_class}
\end{figure}
\begin{table}
	\centering
	\input{figures/final/true_class.tex}
	\caption[True class classification accuracy]{
		\pdfbookmark[1]{Table \ref{tab:true_class}}{true_class}
		\textbf{True class classification accuracy} for several datasets and adjusters.
		After adjustment by the ideal batch adjuster, all true class signal should be preserved, and all classifiers should therefore have the same accuracy in predicting true class before and after adjustment.
		See also Figure \ref{fig:true_class}.
	}
	\label{tab:true_class}
\end{table}

@@

\textbf{What limitations does Confounded have?}
\begin{enumerate*}[(a)]
	\item Confounded uses a variational autoencoder, which are known for often outputting a blurry version of the input data (as can be seen in Figure \ref{fig:mnist}).
	However, recent work has identified modifications that may be made to the basic VAE structure to make output images sharper and more realistic \citep{hou_deep_2016}.
	Similar research with variational autoencoders and gene expression data may yield improved reconstruction losses and decrease the blurring effect.
	\item Confounded takes a long time to run in comparison with ComBat and other linear adjusters.
	Although we acknowledge this as a downfall of many types of machine learning and of neural network in particular, we believe that 30-60 minutes is a reasonable amount of time for a step that will be run only once per pipeline and that can greatly improve data quality.
	\item It can be difficult to identify the optimal network structure and parameter set for a neural network.
	Though this is the case for many applications of neural networks, we feel that Confounded's default structure worked well in our testing and that it will suffice for most batch correction applications.
	\item Neural networks usually perform better when given large amounts of data, and traditional batch datasets typically have very few samples.
	We did find that ComBat may outperform Confounded on many smaller, more traditional datasets, but that Confounded performs better on larger datasets.
	However, Confounded did perform reasonably well with the smaller datasets and appeared to avoid overfitting.
	In cases where ComBat is unable to completely remove batch effects in a small dataset, Confounded may be a viable replacement method.
\end{enumerate*}

\textbf{What else might Confounded be used for?}
At its root, batch correction is a data integration problem:
data from multiple batches must have batch-specific confounding effects removed in order to be treated as one dataset.
Confounded shows promise in removing traditional batch effects from microarray expression data in the Bladderbatch and GSE37199 datasets.
It also effectively decreased artificial batch effects in image data and tissue-specific effects in RNA-Seq data.
Confounded may be effective in other data integration problems, such as combining microarray with RNA-Seq datasets, or merging several large datasets measured under different conditions.

>>>>

% Title / abstract page
\begin{titlingpage}
	\setcounter{page}{2}
	\phantomsection
	\addcontentsline{toc}{section}{Abstract}
	\begin{center}
		ABSTRACT\\
		\vspace{1em}
		\begin{adjustwidth}{+0.5in}{+0.5in}
			\begin{center}
				\thetitle\\
			\end{center}
		\end{adjustwidth}
		\vspace{1em}
		\theauthor\\
		Department of Biology, BYU\\
		Master of Science
	\end{center}
	% \begin{abstract}
		Gene-expression profiling enables researchers to quantify transcription levels in cells, thus providing insight into functional mechanisms of diseases and other biological processes.
		However, because of the high dimensionality of these data and the sensitivity of measuring equipment, expression data often contains unwanted confounding effects that can skew analysis.
		For example, collecting data in multiple runs causes nontrivial differences in the data (known as batch effects),
		known covariates that are not of interest to the study may have strong effects,
		and there may be large systemic effects when integrating multiple expression datasets.
		Additionally, many of these confounding effects represent higher-order interactions that may not be removable using existing techniques that identify linear patterns.
		We created Confounded to remove these effects from expression data.
		Confounded is an adversarial variational autoencoder that removes confounding effects while minimizing the amount of change to the input data.
		We tested the model on artificially constructed data and commonly used gene expression datasets and compared against other common batch adjustment algorithms.
		We also applied the model to remove cancer-type-specific signal from a pan-cancer expression dataset.
		Our software is publicly available at \url{https://github.com/jdayton3/Confounded}.
	% \end{abstract}
	\vfill
	\noindent
	Keywords: batch effects, batch correction, gene expression, transcriptomics, deep learning, adversarial neural network, variational autoencoder
\end{titlingpage}

@@

Gene expression data can be applied in many ways to advance our understanding of medicine and biology.
For example, expression data has been applied to discover conserved genetic modules \citep{stuart_gene-coexpression_2003}, to better understand the mechanisms of cardiovascular disease \citep{henriksen_application_2002}, to more accurately predict the clinical outcome in cancer patients \citep{veer_gene_2002}, and to discover effective drugs for treating specific diseases \citep{sirota_discovery_2011}.
Expression datasets are quite ``wide,'' often containing tens of thousands of columns representing each gene in the human transcriptome.
Because of the sensitive nature of these data (i.e. gene product levels in organisms may respond drastically to small environmental changes) and of the equipment used to quantify expression levels, expression data often contains unwanted confounding effects that can skew analysis.
Three examples of this include
\begin{enumerate*}[(a)]
	\item batch effects, \label{it:batch}
	\item known covariates, and \label{it:cov}
	\item dataset-level effects. \label{it:int}
\end{enumerate*}
\ref{it:batch} Batch effects occur when expression data are generated in multiple runs or multiple batches, and slight systemic differences occur during the different runs, such as different technicians operating the machine or slight temperature differences in the room.
Batch effects are understood to have a nontrivial impact on high-throughput expression data \citep{leek_tackling_2010}.
In one study, researchers found that, contrary to previous knowledge, expression values from mice and humans clustered more closely by species than by tissue type \citep{yue_comparative_2014};
however, referees showed in a rebuttal that when accounting for batch effects, these data actually clustered more closely by tissue type, as initially expected \citep{gilad_reanalysis_2015}.
\ref{it:cov} Systemic bias can be even more pronounced when there are known covariates---for example, using data collected from different cancer types when cancer type is not of interest to the study \citep{dayton_classifying_2017-1}.
\ref{it:int} Finally, though batch effects within a dataset collected all by one lab at one time may be somewhat decreased experimentally by careful replication of experimental conditions, this is no longer possible when integrating different datasets;
however, much better statistical significance can be achieved when analyzing larger datasets, so data integration is a key goal for transcriptomic analysis  \citep{lazar_batch_2013}.
Each of these problems---batch effects, known covariates, and dataset integration---represents a case where data measurements are effected by some hidden variables and must be removed for effective analysis (see Figure \ref{fig:workflow}).

@@

Several methods exist for removing batch effects from gene expression datasets.
Two commonly used methods are ComBat \citep{johnson_adjusting_2007} and SVA \citep{leek_capturing_2007}.
ComBat uses an empirical Bayes method to estimate batch effect parameters and then uses linear regression to remove the effects, and SVA uses singular value decomposition to model batch effects which can then be accounted for in statistical analyses.
Since both of these methods use linear methods to model confounding effects, they are not designed to account for nonlinear effects, such as the cascading changes in gene pathways in response to environmental stressors (e.g. gene X transcripts increase and gene Y transcripts decrease in response to rising temperature, and since both are now past a certain threshold, gene Z transcripts also increase, and so on).
As machine learning becomes more common in biological research, these nonlinear confounding effects become more troublesome since many machine learning algorithms can successfully identify complex interactions between variables.
For example, advances in neural networks have introduced new ways to account for higher-order, nonlinear relationships in data.
These networks have proven effective in removing irrelevant, domain-specific signal in credit rating, online reviews, and image recognition tasks \citep{louizos_variational_2015}.
Recent work \cite{shaham_removal_2017,shaham_batch_2018,upadhyay_removal_2019} has applied neural networks to batch effects.
However, several limitations complicate their usability on many real-world datasets:
they require that the input data only contains two batches, that the batches are sufficiently large (we received errors when testing with a subset of MNIST where n=100), and that the batches are perfectly balanced.
These requirements rarely hold in existing datasets; for example, the bladderbatch dataset used in the R sva package \cite{leek_bladderbatch_2017,leek_sva_2017} has 5 batches, only 57 samples, and between 4 and 19 samples per batch.
Additionally, the metrics these studies have used to validate results don't test whether complex interactions still remain in the data.

@@

In this study, we present Confounded, an adversarial autoencoder that identifies and removes confounding effects.
We test the hypothesis that using an adversarial neural network can correct for confounding effects more completely than previous tools do.
We also explore the extent to which confounders still remain in different datasets after adjustment with various algorithms, and we present a framework to assess the extent to which confounding effects remain after adjustment using various classification algorithms.

@@

We used an adversarial autoencoder network to model and remove the confounding effects.
We structured this network in two parts: a variational autoencoder \cite{louizos_variational_2015} to replicate the input (expression) data and a discriminator to detect remaining confounding effects in the autoencoder's output.
By penalizing the autoencoder for the discriminator's success, the autoencoder subnetwork learned over the course of training to output the expression data with confounding effects minimized.
We implemented the neural network in TensorFlow 1.11.0 \cite{abadi_tensorflow_2015} with Python 3.6 \cite{python_software_foundation_python_2019}.
All layers in the network were fully connected and all activation functions were Rectified Linear Units (ReLUs) \cite{agarap_deep_2018} except the final layers in the autoencoder and the discriminator, which used the sigmoid function.

@@

We implemented the variational autoencoder \citep{louizos_variational_2015} described by \citet[Chapter 15]{geron_hands-machine_2017}.
This network has 2 hidden encoding layers and 2 decoding layers, each of size 500.
The code size is 20.
Each hidden layer is activated with the Exponential Linear Unit (ELU) function \citep{clevert_fast_2015}.
It is trained with the Adam optimizer \citep{kingma_adam_2014} on reconstruction loss (sigmoid cross entropy) combined with latent loss (KL divergence \citep{kullback_information_1951}).

@@

We trained the network using three loss functions.
First, we calculated the autoencoder's loss ($L_A$) by summing the reconstruction loss (sigmoid cross entropy between the autoencoder's input and output) and the latent loss (Kullback-Leibler (KL) divergence \cite{kullback_information_1951} of the code layer).
Second, we calculated the discriminator's loss ($L_D$) as sigmoid cross entropy between its output and a one-hot encoding of the samples' batch labels.
Finally, we also trained the autoencoder layers on a combination of the two previous losses,

@@

The $\lambda$ value represents a tradeoff parameter for tuning the network's tendency for more faithfully replicating the input or for more completely removing confounding effects.
A higher $\lambda$ value indicates that the network should remove confounding effects more aggressively, whereas a lower value indicates that the network should instead favor faithfully reconstructing the input data.
We did not optimize $L_A$ directly; instead we trained the autoencoder by optimizing $L_{dual}$.

@@

In all cases, we trained the network using the Adam Optimizer \citep{kingma_adam_2014} with a training rate of 0.0001 for 10,000 iterations on mini-batches of size 100.
In each iteration, we optimized on both $L_D$ and $L_{dual}$.
When optimizing $L_D$ (i.e. training the discriminator), we froze the autoencoder's weights, and vice versa.
We trained the network on a 2017 Dell XPS 15 9560 with an 8-core Intel i7-7700HQ CPU and 16 GB of RAM.
For each dataset, training typically took roughly 30 minutes to complete, including the time taken to load the input into memory and to save the output to disk.

\begin{figure}
	\pdfbookmark[1]{Figure \ref{fig:training_loss}}{training_loss}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/training_loss.pdf}
	\caption[Autoencoder and discriminator loss]{\textbf{Autoencoder and discriminator loss} over time for one run of Confounded on the MNIST dataset.
	Over the course of training, the autoencoder more faithfully replicates the input data.
	The autoencoder also seems to introduce noise (see the red dashed line around iteration 3100) in response to the discriminator's slight improvements.}
	\label{fig:training_loss}
\end{figure}

@@

Because there is no batch information in the MNIST digits dataset, we had to simulate nonlinear confonding effects.
To do so, we wrote a Python script to take the MNIST data in, apply a nonlinear effect, and output the adjusted data.
We applied nonlinear effects by iteratively realizing vectors of normally distributed values, multiplying and adding these vectors to the ``expression'' vectors, and applying nonlinearity to the adjusted vector.
We split the image data into two batches while keeping the batches balanced (5,000 images for each batch) and including the same number of each digit in each batch.
We applied the same random vectors to each image in a batch.
Finally, we added random noise to each image in order to prevent images in a batch from being overly similar to each other.

@@

The GSE37199 dataset contains Affymetrix microarray gene-expression data from patients with advanced castration-resistant prostate cancer \cite{olmos_prognostic_2012}.
We accessed a version of this dataset from \url{http://doi.org/10.17605/OSF.IO/SSK3T} that was tidied as part of a curated compendium of human transcriptional biomarker data \cite{golightly_curated_2018}.
It contains expression values for 93 tissue samples categorized as either ``Advanced castration resistant'' or ``good prognosis.''
We used this cancer status variable as the ``true class'' for classification.
It has two types of batch variables: ``plate'' and ``centre.''
We adjusted against the ``plate'' variable because it was more balanced than ``centre'' (with counts of $\{43, 50\}$ compared to $\{27, 66\}$).
The GSE37199 dataset represents a slightly larger dataset than bladderbatch, with only two batches that are closer to being balanced (with batch counts of $\{4, 5, 11, 18, 19\}$ and $\{43, 50\}$, respectively).

@@

We compared our method to two other batch adjusters: a scale adjuster and ComBat \citep{johnson_adjusting_2007}.
We implemented the scale adjuster in the R programming language, version 3.6.0 \citep{r_core_team_r_2014} using RStudio version 1.2.1194 \citep{rstudio_team_rstudio_2018}.
It adjusts the data by linearly expanding or contracting each batch so all batches have the same range.
We used the ComBat implementation from the R sva package \citep{leek_sva_2017} with some modifications to allow it to work on columns without variance in the MNIST dataset.
We initially intended to test against the SVA \citep{leek_capturing_2007} method but concluded that SVA is more suited for producing surrogate variables for further statistical research rather than removing those variables from the data.
There are a number of other methods for batch adjustment that do and do not use deep neural networks (for example, see \citep{leek_capturing_2007,espin-perez_comparison_2018,shaham_removal_2017,shaham_batch_2018}). % TODO: add like 8 more methods
Unfortunately, most of these methods lack a common interface and common assumptions that input datasets must meet.
For these reasons, we compared Confounded only to the two methods listed above.
Future batch adjustment research may benefit from standardization of input formats, user interfaces, and validation datasets.

@@

The scaling method performed consistently worse across the qualitative and quantitative evaluations that we performed;
ComBat and Confounded performed relatively well overall, but each of these algorithms excelled in different types of scenarios, which we illustrate below.

\subsection{Confounded removes nonlinear confounding effects that other adjusters miss}

Various previous studies use differing methods to show qualitatively and quantitatively how well confounding effects have been removed.
To compare Confounded to other batch adjustment methods, we compared PCA and t-SNE plots along with MSE, MMD, and classifier batch prediction accuracy (using various classifiers).

@@

Confounded shows mixed success with the MSE and MMD metrics.
With MSE, Confounded outperformed the scale adjuster in 3 of the 4 datasets but scored drastically worse on the MNIST dataset, with scores (listed as: scale, Confounded) of \{(0.003, 0.00168), (0.247, 0.0698), (0.00312, 0.0187), (912315, 139458)\}.
With MMD, Confounded outperformed the scale adjuster again in 3 of the 4 datasets and tied the scale adjuster on the TCGA dataset, with scores of \{(0.463, 0.258),(0.0906, 0.0376),(0.0653, 0.0117),(0.0942, 0.0942)\}.
With both metrics, Confounded consistently performed somewhat worse than ComBat.
(See Figures \ref{fig:mse} and \ref{fig:mmd}.)

@@

We would expect that after batch adjustment by an ideal adjuster, batch would no longer be detectable by any machine learning classifier.
Using the batch classification accuracy metric, Confounded seems to outperform other adjusters on larger datasets, whereas ComBat and Confounded seem to perform about the same on smaller datasets (see Figure \ref{fig:batch}).
With both the bladderbatch and GSE37199 datasets, batch classification accuracy decreases well below baseline after batch adjustment with ComBat for all classifiers we tested (see Table \ref{tab:batch}).
Interestingly, batch accuracy also decreases drastically for the MNIST and TCGA datasets, but only for the Naive Bayes classifier.
This may be due to two factors: both ComBat and Naive Bayes use Bayesian methods, so ComBat may specifically remove the effects that Naive Bayes identifies; and Naive Bayes does not find patterns based on interactions between variables.
Although Naive Bayes is no longer able to identify confounding effects in the data after ComBat-adjustment, Random Forests (which does use interactions between variables) still has a very high accuracy for MNIST and an increased accuracy for TCGA.
In contrast, after adjustment by Confounded, the Random Forests algorithm's accuracy decreases more than with any other adjuster for both the MNIST and TCGA datasets.
This indicates that while ComBat's performance may work at least as well as Confounded for smaller expression datasets, Confounded may work better with larger datasets.

@@

With the larger datasets, Confounded's true class accuracy consistently decreases below the accuracy of other adjusters.
A look at the MNIST digits before and after adjustment (see Figure \ref{fig:mnist}) shows that Confounded's output is often blurry, as is common with the output of variational autoencoders \cite{hou_deep_2016}.
With MNIST, Confounded's accuracy with Random Forests is still much higher than baseline (84.8\% vs. 11.4\%), but with TCGA, the accuracy decreases below baseline (66.5\% vs. 69.8\%) while the other adjusters' accuracies remain above baseline.
However, the particular set of parameters that we used in Confounded are likely not optimal for every dataset.
Additional tuning may improve the performance metrics.

\begin{figure}
	\pdfbookmark[1]{Figure \ref{fig:true_class}}{true_class}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/true_class_accuracy.pdf}
	\caption[True class classification accuracy]{\textbf{True class classification accuracy} for several datasets and adjusters with 4-fold cross-validation repeated 3 times. A higher accuracy after adjustment is desired because it represents that the adjuster has not destroyed the true class signal.
	See also Table \ref{tab:true_class}.}
	\label{fig:true_class}
\end{figure}
\begin{table}
	\centering
	\input{figures/final/true_class.tex}
	\caption[True class classification accuracy]{
		\pdfbookmark[1]{Table \ref{tab:true_class}}{true_class}
		\textbf{True class classification accuracy} for several datasets and adjusters.
		After adjustment by the ideal batch adjuster, all true class signal should be preserved, and all classifiers should therefore have the same accuracy in predicting true class before and after adjustment.
		See also Figure \ref{fig:true_class}.
	}
	\label{tab:true_class}
\end{table}

@@

\textbf{What limitations does Confounded have?}
\begin{enumerate*}[(a)]
	\item Confounded uses a variational autoencoder, which are known for often outputting a blurry version of the input data (as can be seen in Figure \ref{fig:mnist}).
	However, recent work has identified modifications that may be made to the basic VAE structure to make output images sharper and more realistic \citep{hou_deep_2016}.
	Similar research with variational autoencoders and gene expression data may yield improved reconstruction losses and decrease the blurring effect.
	\item Confounded takes a long time to run in comparison with ComBat and other linear adjusters.
	Although we acknowledge this as a downfall of many types of machine learning and of neural network in particular, we believe that 30-60 minutes is a reasonable amount of time for a step that will be run only once per pipeline and that can greatly improve data quality.
	\item It can be difficult to identify the optimal network structure and parameter set for a neural network.
	Though this is the case for many applications of neural networks, we feel that Confounded's default structure worked well in our testing and that it will suffice for most batch correction applications.
	\item Neural networks usually perform better when given large amounts of data, and traditional batch datasets typically have very few samples.
	We did find that ComBat may outperform Confounded on many smaller, more traditional datasets, but that Confounded performs better on larger datasets.
	However, Confounded did perform reasonably well with the smaller datasets and appeared to avoid overfitting.
	In cases where ComBat is unable to completely remove confounding effects in a small dataset, Confounded may be a viable replacement method.
\end{enumerate*}

\textbf{What else might Confounded be used for?}
At its root, batch correction is a data integration problem:
data from multiple batches must have batch-specific confounding effects removed in order to be treated as one dataset.
Confounded shows promise in removing traditional batch effects from microarray expression data in the Bladderbatch and GSE37199 datasets.
It also effectively decreased artificial batch effects in image data and cancer-type-specific confounding effects in RNA-Seq data.
Confounded may be effective in other data integration problems, such as combining microarray with RNA-Seq datasets, or merging several large datasets measured under different conditions.
